Normalization Circularity in Population-Based FA Analysis

### 2011-06-07 ANTsTBSS/paper.tex
Minor edits.

\begin{abstract}
Template-based normalization plays a critical role in many population
studies based on diffusion tensor imaging (DTI).  Several current
approaches compute the fractional anisotropy (FA) from the DTI and then
directly register the FA image to a FA template before performing
statistical comparison between different groups within a population.  In
this paper, we demonstrate that this analysis strategy induces selection
bias that varies in strength with the similarity metric used in
registration.  Specifically, we show that the popular sum of squared
difference intensity metric maximizes effect size, rather than
anatomical alignment, and consistently overestimates statistical
significance.  Consequently, we advocate a more traditional approach
that normalizes a population of FA images through the associated T1
anatomical images.  We show that the Symmetric Group Normalization
strategy for unbiased template construction from anatomical (e.g.
T1-weighted) images provides a valid and unbiased coordinate system for
statistical inference of FA differences and yet retains detection power
in comparison to FA-based normalization methods.  We establish this
outcome by using open data in normal controls from the International
Neuroimaging Data-sharing Initiative as well as a cohort of 16 traumatic
brain injury survivors compared with 17 control subjects.  Alignment
comparisons with the normalization component of the popular tract-based
spatial statistics (TBSS) show that our anatomically-based template
approach produces good alignments despite the absence of any direct
FA-to-FA registration.  Of practical consideration we note that the
software tools discussed in this work are provided to the public as part
of the open source Advanced Normalization Tools (ANTs) repository.  In conclusion, we recommend against circular analysis strategies that select image similarity metrics that are tied directly to the ultimate statistical functions used to assess population effects.

@@

The seminal work of Basser et al. \cite{Basser1994a,Basser1994} established diffusion tensor imaging (DTI) as a viable investigatory MRI technique.  DTI's sensitivity to brain architecture \cite{Basser1996,Assaf2008} enables multiple uses in assessing neuro-structural differences in cross-population studies \cite{Kubicki2005,Arnone2008,Kantarci2010,Rametti2010} including subjects with traumatic brain injury (TBI) \cite{Caeyenberghs2010,Jiang2010,Warner2010}.  Two popular comprehensive software packages for assessing population FA differences include the SPM Matlab-based toolkit%
\footnote{
http://www.fil.ion.ucl.ac.uk/spm/
}
for voxel-based morphometry (VBM)
and the tract-based spatial statistics (TBSS) framework \cite{Smith2006}.%
\footnote{
http://www.fmrib.ox.ac.uk/fsl/tbss/index.html
} 
VBM\cite{Ashburner2001} continues to find application to FA studies\cite{Kakeda2010,Takao2010} despite concerns about its validity\cite{Bookstein2001,Davatzikos2004} in DTI \cite{Jones2005,Chung2008}.  The TBSS framework was partially developed in response to these concerns in which voxel values are projected onto a template white matter skeleton for increased statistical power.  
%(e.g. \cite{Arnone2008,Bodini2009}).

Although there are substantive differences in normalization schemes
between these and other image registration approaches, a core
commonality includes direct FA-to-FA registration.  The sum of squared differences (SSD) image similarity metric is perhaps the easiest to interpret (it drives the image difference to zero), is computationally efficient (just compute the image difference and an image gradient) and is therefore widely used.  Some of the most popular registration methods (SPM, Demons, FSL's FNIRT) rely on this metric and provide reasonable performance levels \cite{Klein2009}.
\footnote{
In addition to TBSS and SPM, other popular packages in the neuroimaging community 
which use the SSD metric includes AIR \cite{Woods1998} and LDDMM \cite{Beg2005}.
}  
Now, suppose the SSD metric is optimized to find a set of $M$ transformations that map a population of $M$ FA images, $\{I_1, I_2, \ldots, I_M\}$, to a representative FA template, $J$. Then, the algorithm will compute transformations
$\boldsymbol{\mathcal{T}} = \{\mathcal{T}_1, \mathcal{T}_2,\ldots,\mathcal{T}_M \}$, which minimize the SSD metric over the population, i.e.
\begin{align}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{m=1}^M\sum_{n=1}^N \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2.
\end{align}
Switching the order of the summations (since the transformations are mutually independent) and recognizing that a principal criterion for the selection of $J$ is such that it be a good approximation of the mean of the aligned images, it becomes apparent that the transformation solution
\begin{align}\label{eq:variance}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{n=1}^N \underbrace{\sum_{m=1}^M \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2}_{
    \propto\mathrm{\,voxelwise\,variance}}
\end{align}
is that which minimizes the voxelwise variance.  Reducing variance will also directly affect the value of a standard population statistic, the Student's $t$-test.  Other standard similarity metrics will also exhibit this
trend to some degree as they minimize some local measure of intensity-based distance between the subject and the template.

Thus, instead of normalization based on anatomical alignment followed by
statistical testing, direct FA normalization explicitly conflates
anatomical alignment with optimizing the very statistical testing
results that one will ultimately use to assess hypotheses.  This immediately evokes a sense of circularity in the analysis \cite{Kriegeskorte2010}.
According to Kriegeskrote et al., ``An analysis is circular (or nonindependent) if it is based on data that were selected for showing the effect of interest or a related effect.''  As illustrated with
Eqn. \ref{eq:variance}, direct FA-to-FA registration will find transformations that align (or select) voxels such that the statistical testing result (or the effect of interest) is increased (assuming one is using Student's test).  While the SSD metric is able to provide visually attractive alignments across subjects, the voxel-level details (that is, which voxels are matched where) are, as we show in the following simulation study, strikingly important in terms of the effect on detection power.

@@

After creation of the simulated data set, we performed voxelwise testing (both $t$-testing
and Wilcoxon rank-sum testing, one-tailed, $\alpha = 0.95$) with false discovery
rate correction within the white matter to get ``true'' cumulative distribution functions (CDFs).  The CDF vertical axis shows the cumulative number of voxels with corrected p-values at or below a given significance level (horizontal axis) and thus may be used as a surrogate measure for spatial sensitivity.  From here, we denote these graphs as "statistical significance curves."  These curves are plotted in black in Fig. \ref{fig:simulated_testing}.  To illustrate the 
effect of similarity metric on statistical significance, we performed ANTs registration \cite{Avants2011} of each subject to the template
(Fig. \ref{fig:simulated_data}(b)) using cross correlation (CC), mutual information (MI), 
Demons \cite{Thirion1998}, and sum of squared differences (SSD). A representative deformed image
using the SSD metric and its corresponding deformed grid is given in 
Fig. \ref{fig:simulated_data_warped}. 
The resulting statistical significance curves are plotted with the true significance curves in Fig. 
\ref{fig:simulated_testing}.

This example demonstrates the undesirable phenomenon previously described.  Even though these images were anatomically aligned prior to registration, the resulting deformations (falsely) increased detection power.  This effect can be explained in the context of Eqn. (\ref{eq:variance}) and by the fact that we added noise to the underlying distribution which allows the deformable mapping algorithms the freedom to transform the positions of the most "powerful" voxels to minimize variance across the population (and increase the t-test).  Note, however, that this effect is increased by the fact that noise was added to the data. {\textcolor{red}{FIXME --- does more / less noise increase / decrease the exaggerated power? }}
Common similarity metric functions can therefore induce false positives by minimizing local intensity 
variance between the subjects and template which explicitly maximizes statistical
testing results in contrast to maximizing anatomical alignment.  This effect is strongest in 
the SSD and Demons metrics and weakest in the MI and CC metrics likely due
to the degree of correspondence to Eqn. (\ref{eq:variance}).   This is also supported by contrasting
the $t$-test results given in Fig. \ref{fig:simulated_testing}(a) with the non-parametric Wilcoxon
rank-sum testing (Fig. \ref{fig:simulated_testing}(b)).  The rank test does not explicitly use variance in its assessment and thus exhibits far less statistical exaggeration.

@@

Population-specific templates (or optimal templates) alleviate some of
these issues by deriving a most representative image from the population
\cite{FIXME}.  Large deformation registration algorithms also reduce
this confound by being less sensitive to the deformation distance
between subject and target.  Some approaches combine both advantages,
for instance, the diffeomorphic approach of Joshi et al. who employ the
SSD metric and a shape distance to bring the subject group of images
into alignment \cite{Joshi2004}.  Variants
include extension to multiple modalities \cite{Lorenzen2006} and small deformations
\cite{Geng2009}.  These approaches iteratively minimize group difference in ``congealing''
towards a representative image template \cite{Learned-Miller2006}.

@@

The CDF of the voxelwise FA coefficient of variation {\textcolor{red}{ FIXME define coefficient of variation and explain what it means --- how does it relate to variance?}} of the aligned
images is shown for each case in Figure \ref{fig:tbi_cov}.  This is
very much related to earlier findings in that strategies evincing
greater alignment, as quantified by the coefficient of variation,
will trend towards smaller variance thus exhibiting a left-shifted
CDF.  Although direct FA-to-FA registration using ANTs with common
similarity metrics gives ``better alignment'' such alignment is 
produced at a cost of significant type 1 errors.  Common subsampling and
smoothing strategies and local neighborhood metrics such as CC and MI reduce the 
Type 1 error problem (as evidenced in Fig. \ref{fig:tbi_testing}) 
but also show a decrease in alignment.  This is further supported
by the overlap measures \cite{Crum2006} given in Table \ref{table:lom}.
For each alignment strategy, we thresholded the derived mean FA image (FA $
\geq$ 0.2) and compared it with white matter mask of the MNI152 template
segmented using Atropos \cite{Avants2011a}.  The template-based
strategy showed the best overall overlap with the white matter MNI152
segmentation.

@@

\subsubsection{Statistical Power Related to Gender-based FA Differences}

@@

\subsection{Discussion} Prior evaluation work using anatomically labeled
data \cite{bunch_of_stuff} indicates that mutual information,
cross-correlation and the SSD/Demons metric are all capable of producing
high quality anatomical alignment.  All of this work, though, compares
the quality of alignment at a relatively coarse scale of major gyri,
lobes and regions.  At the voxel level, it is nearly impossible to
determine the ground truth correspondence.  Indeed, there is no reason
to believe that subtle differences between a set of alignment strategies
(induced, in the experiments of this paper, by changing the similarity
metrics used in image registration) that all apparently work well should
be considered significant.  This work, however, highlights a dramatic impact of similarity metric choice on detection power in template-based FA studies.  Our contention is that this dramatic "improvement" in detection power is not due to better anatomical alignment.  Rather, it is a result of circularity/non-independence of the normalization and statistical estimation strategy.  Consequently, we recommend that future work should take greater care when pairing the normalization strategy with the statistical framework.  Furthermore, we recommend that researchers should diligently seek to maintain independence between these two critical stages of a population study.  That is, {\textcolor{red}{{\em the features that drive the minimization of a similarity metric should be independent of the features that will be used in hypothesis testing.)}}}

>>>>

\begin{abstract}
Template-based normalization plays a critical role in many population
studies based on diffusion tensor imaging (DTI).  Several current
approaches compute the fractional anisotropy (FA) from the DTI and then
directly register the FA image to a FA template before performing
statistical comparison between different groups within a population.  In
this paper, we demonstrate that this analysis strategy induces selection
bias that varies in strength with the similarity metric used in
registration.  Specifically, we show that the popular sum of squared
difference intensity metric maximizes effect size, rather than
anatomical alignment, and consistently overestimates statistical
significance.  Consequently, we advocate a more traditional approach
that normalizes a population of FA images through the associated T1
anatomical images.  We show that the Symmetric Group Normalization
strategy for unbiased template construction from anatomical (e.g.
T1-weighted) images provides a valid and unbiased coordinate system for
statistical inference of FA differences and yet retains detection power
in comparison to FA-based normalization methods.  We establish this
outcome by using open data comprised of normal controls from the International
Neuroimaging Data-sharing Initiative as well as a cohort of 16 traumatic
brain injury survivors compared with 17 control subjects.  Alignment
comparisons with the normalization component of the popular tract-based
spatial statistics (TBSS) show that our anatomically-based template
approach produces good alignments despite the absence of any direct
FA-to-FA registration.  Of practical consideration we note that the
software tools discussed in this work are provided to the public as part
of the open source Advanced Normalization Tools (ANTs) repository.  In conclusion, we recommend against circular analysis strategies that select image similarity metrics that are tied directly to the ultimate statistical functions used to assess population effects.

@@

The seminal work of Basser et al. \cite{Basser1994a,Basser1994} established diffusion tensor imaging (DTI) as a viable investigatory MRI technique.  DTI's sensitivity to brain architecture \cite{Basser1996,Assaf2008} enables multiple uses in assessing neuro-structural differences in cross-population studies \cite{Kubicki2005,Arnone2008,Kantarci2010,Rametti2010} including subjects with traumatic brain injury (TBI) \cite{Caeyenberghs2010,Jiang2010,Warner2010}.  Two popular comprehensive software packages for assessing population FA differences include the SPM Matlab-based toolkit%
\footnote{
http://www.fil.ion.ucl.ac.uk/spm/
}
for voxel-based morphometry (VBM)
and the tract-based spatial statistics (TBSS) framework \cite{Smith2006}.%
\footnote{
http://www.fmrib.ox.ac.uk/fsl/tbss/index.html
} 
VBM \cite{Ashburner2001} continues to find application to FA studies \cite{Kakeda2010,Takao2010} despite concerns about its validity \cite{Bookstein2001,Davatzikos2004} in DTI \cite{Jones2005,Chung2008}.  The TBSS framework was partially developed in response to these concerns in which voxel values are projected onto a template white matter skeleton for increased statistical power.  
%(e.g. \cite{Arnone2008,Bodini2009}).

Although there are substantive differences in normalization schemes
between these and other image registration approaches, a core
commonality includes direct FA-to-FA registration.  The sum of squared differences (SSD) image similarity metric is perhaps the easiest to interpret (it drives the image difference to zero), is computationally efficient (just compute the image difference and an image gradient) and is therefore widely used.  Some of the most popular registration methods (SPM, Demons, FSL's FNIRT) rely on this metric and produce reasonable performance levels \cite{Klein2009}.%
\footnote{
In addition to TBSS and SPM, other popular packages in the neuroimaging community 
which use the SSD metric include AIR \cite{Woods1998} and LDDMM \cite{Beg2005}.
}

However, suppose the SSD metric is optimized to find a set of $M$ transformations that map a population of $M$ FA images, $\{I_1, I_2, \ldots, I_M\}$, to a representative FA template, $J$. The algorithm will then compute the transformations,
$\boldsymbol{\mathcal{T}} = \{\mathcal{T}_1, \mathcal{T}_2,\ldots,\mathcal{T}_M \}$, which minimize the SSD metric over the population, i.e.
\begin{align}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{m=1}^M\sum_{n=1}^N \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2.
\end{align}
Switching the order of the summations (since the transformations are mutually independent) and recognizing that a principal criterion for the selection of $J$ is such that it be a good approximation of the mean of the aligned images, it becomes apparent that the transformation solution
\begin{align}\label{eq:variance}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{n=1}^N \underbrace{\sum_{m=1}^M \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2}_{
    \propto\mathrm{\,voxelwise\,variance}}
\end{align}
is that which minimizes the voxelwise variance.  Reducing variance will also directly affect the value of a standard population statistic, the Student's $t$-test.  Other standard similarity metrics will also exhibit this
trend to some degree as they minimize some local measure of intensity-based distance between the subject and the template.

Thus, instead of normalization based on anatomical alignment followed by
statistical testing, direct FA normalization explicitly conflates
anatomical alignment with optimizing the very statistical testing
results that one will ultimately use to assess hypotheses.  This immediately evokes a sense of circularity in the analysis \cite{Kriegeskorte2010}.
According to Kriegeskrote et al., ``An analysis is circular (or nonindependent) if it is based on data that were selected for showing the effect of interest or a related effect.''  As illustrated with
Eqn. \ref{eq:variance}, direct FA-to-FA registration will find transformations that align (or select) voxels such that the statistical testing result (or the effect of interest) is increased (assuming one is using Student's 
$t$-test).  While the SSD metric is able to provide visually attractive alignments across subjects, the voxel-level details (that is, which voxels are matched where) are, as we show in the following simulation study, strikingly important in terms of the effect on detection power.

@@

After creation of the simulated data set, we performed voxelwise testing (both $t$-testing
and Wilcoxon rank-sum testing, one-tailed, $\alpha = 0.95$) with false discovery
rate correction within the white matter to get ``true'' cumulative distribution functions (CDFs).  The CDF vertical axis shows the cumulative number of voxels with corrected $p$-values at or below a given significance level (horizontal axis) and thus may be used as a surrogate measure for spatial sensitivity.  From here, we denote these graphs as "statistical significance curves."  These curves are plotted in black in Fig. \ref{fig:simulated_testing}.  To illustrate the 
effect of similarity metric on statistical significance, we performed ANTs registration \cite{Avants2011} of each subject to the template
(Fig. \ref{fig:simulated_data}(b)) using cross correlation (CC), mutual information (MI), 
Demons \cite{Thirion1998}, and sum of squared differences (SSD). A representative deformed image
using the SSD metric and its corresponding deformed grid is given in 
Fig. \ref{fig:simulated_data_warped}. 
The resulting statistical significance curves are plotted with the true significance curves in Fig. 
\ref{fig:simulated_testing}.

This example demonstrates the undesirable phenomenon previously described.  Even though these images were anatomically aligned prior to registration, the resulting deformations (falsely) increased detection power.  This effect can be explained in the context of Eqn. (\ref{eq:variance}) and by the fact that we added noise to the underlying distribution which allows the deformable mapping algorithms the freedom to transform the positions of the most "powerful" voxels to minimize variance across the population (and increase the $t$-test).  Note, however, that this effect is increased by the fact that noise was added to the data.  Decreasing the standard deviation of the noise causes an increase in separation between  the SSD and Demon's metric pair with the MI and CC metric pair with the latter pair closer to the the true significance curve. Increasing the noise yields the same ranking with a more even spread between significance curves.

Common similarity metric functions can therefore induce false positives by minimizing local intensity 
variance between the subjects and template which explicitly maximizes statistical
testing results in contrast to maximizing anatomical alignment.  This effect is strongest in 
the SSD and Demons metrics and weakest in the MI and CC metrics likely due
to the degree of correspondence to Eqn. (\ref{eq:variance}).   This is also supported by contrasting
the $t$-test results given in Fig. \ref{fig:simulated_testing}(a) with the non-parametric Wilcoxon
rank-sum testing (Fig. \ref{fig:simulated_testing}(b)).  The rank test does not explicitly use variance in its assessment and thus exhibits far less statistical exaggeration even though it offers roughly 95\%
the power of the $t$-test assuming certain assumptions are met \cite{Siegel1988,Rorden2007}.

@@

Population-specific templates (or optimal templates) alleviate some of
these issues by deriving a most representative image from the population
\cite{Good2001}.  Large deformation registration algorithms also reduce
this confound by being less sensitive to the deformation distance
between subject and target.  Some approaches combine both advantages,
for instance, the diffeomorphic approach of Joshi et al. who employ the
SSD metric and a shape distance to bring the subject group of images
into alignment \cite{Joshi2004}.  Variants
include extension to multiple modalities \cite{Lorenzen2006} and small deformations
\cite{Geng2009}.  These approaches iteratively minimize group difference in ``congealing''
towards a representative image template \cite{Learned-Miller2006}.

@@

The CDF of the voxelwise FA coefficient of variation of the aligned
images is shown for each case in Figure \ref{fig:tbi_cov}.  
The coefficient of variation at each voxel is defined as the variance
divided by the mean over the population of FA values.  Thus,
those metrics which best minimize the voxelwise variance would
trend towards greater alignment as quantified by the coefficient of
variation.

%This is very much related to earlier findings in that strategies evincing
%greater alignment, as quantified by the coefficient of variation,
%will trend towards smaller variance thus exhibiting a left-shifted
%CDF.

Although direct FA-to-FA registration using ANTs with common
similarity metrics gives ``better alignment'' such alignment is 
produced at a cost of significant type 1 errors.  Common subsampling and
smoothing strategies and local neighborhood metrics such as CC and MI reduce the 
Type 1 error problem (as evidenced in Fig. \ref{fig:tbi_testing}) 
but also show a decrease in alignment.  This is further supported
by the overlap measures \cite{Crum2006} given in Table \ref{table:lom}.
For each alignment strategy, we thresholded the derived mean FA image (FA $
\geq$ 0.2) and compared it with white matter mask of the MNI152 template
segmented using Atropos \cite{Avants2011a}.  The template-based
strategy showed the best overall overlap with the white matter MNI152
segmentation.

@@

\subsubsection{Statistical Power Related to Gender-Based FA Differences}

@@

\section{Discussion} Prior evaluation work using anatomically labeled
data indicates that mutual information,
cross-correlation and the SSD/Demons metric are all capable of producing
high quality anatomical alignment (e.g. \cite{Klein2009}).  All of this work, though, compares
the quality of alignment at a relatively coarse scale of major gyri,
lobes and regions.  At the voxel level, it is nearly impossible to
determine the ground truth correspondence.  Indeed, there is no reason
to believe that subtle differences between a set of alignment strategies
(induced, in the experiments of this paper, by changing the similarity
metrics used in image registration) that all apparently work well should
be considered significant.  This work, however, highlights a dramatic impact of similarity metric choice on detection power in template-based FA studies.  Our contention is that this dramatic "improvement" in detection power is not due to better anatomical alignment.  Rather, it is a result of circularity/nonindependence of the normalization and statistical estimation strategy.  Consequently, we recommend that future work should take greater care when pairing the normalization strategy with the statistical framework.  Furthermore, we recommend that researchers should diligently seek to maintain independence between these two critical stages of a population study.  That is, {the features that drive the minimization of a similarity metric should be independent of the features that will be used in hypothesis testing.  In particular, one should avoid combining the SSD metric for normalizing FA with Student's t-test for FA differences.
